The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-06-06 pkg/partykit/inst/guideliketest/draft/draft.Rnw
small corrections; added information on where to find code to reproduce simulation study (arxiv)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2613 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: abfe84cfdb0c2cf1b1f6c774ee588cb4039acfc1

if(file.exists("tree.rda")){
  load("tree.rda")
} else {
  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
                         nrsplitvar = 2, nrsteps = 1, nrlevels = 2,
                         delta = seq(from = 0, to = 1, by = 0.1),
                         xi = c(0, 0.2, 0.5, 0.8), vary_beta = "all",
                         binary_regressor = FALSE, binary_beta = TRUE,
                         only_intercept = FALSE,
                         test = c("ctree", "mfluc", "ctree_max",
                                  "ctree_cat", "mfluc_cat", "ctree_max_cat",
                                  "ctree_bin", "mfluc_bin", "ctree_max_bin",
                                  "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
                         beta0 = NULL, beta1 = NULL,
                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05)

  save(simres, file = "tree.rda")
}
sim_tree <- simres$res
rm(simres)

@@

The top row of Figure~\ref{fig:tree_all} presents the results for pre-pruning which, not surprisingly,
reflect the results from previous sections. Thus, CTree and MOB perform similarly and rather well
as the underlying significance tests have good power properties. Only for late abrupt shifts the
linear test statistics in CTree clearly have lower power than the maximally-selected statistics in MOB.
Again, this picture would reverse for smooth changes instead of abrupt shifts. Compared to CTree/MOB
both GUIDE flavors clearly perform worse as they underlying tests are less powerful; notably
for the residual-based GUIDE which again has problems picking up the slope change associated with $Z_2$.

@@

When switching from pre-pruning to post-pruning (bottom row of Figure~\ref{fig:tree_all})
it is indeed shown that many of the problems stemming from the low power of the two GUIDE
flavors are indeed mitigated. Thus, by first growing a large tree without stopping upon
non-significance and then pruning back based on predictive performance instead substantially
improves the fit of the two GUIDE flavors. However, using the residuals only in GUIDE still
performs clearly worse compared to the GUIDE+scores. The latter is essentially on par with
CTree and MOB when the true split matches one of the categorization bins (i.e., $xi = 0$
and $\xi = 0.5$, respectively) while CTree/MOB still perform somewhat better in the other
cases ($\xi = 0.2$ and $\xi = 0.8$).

In summary, there is clear support for the conventional wisdom that the power of the testing
strategy in an unbiased tree algorithm is not so important when combined with post-pruning.
However, there are limits to this. Consequently, when tests can be made more powerful --
e.g., by using scores instead of residuals -- then this improves the corresponding tree
algorithm. Finally, the simulation also supports using pre-pruning in an unbiased tree algorithm
when the underlying testing strategy that also works well as a classical significance test.

@@

Linear models as presented in this study have been chosen as they are highly relevant in practice,
allow for simple illustrations, and theoretical insights are available for the testing strategies.
However, the results can be easily extended to a wide variety of other models where the introduced building
blocks can be applied in the same way. Specifically, it has been shown theoretically that certain
some changes in the parameters do not lead to shifts in the residuals (but in other components of the
scores). \cite{Ploberger+Kraemer:1992} showed that residual-based tests can detect a change in the parameters
of a linear model only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor which in our case is $(1,0)^{\top}$.
Consequently, if only the slope $\beta_1$ but not the intercept $\beta_0$ changes, the shift
is of type $(0, \delta)^\top$ and thus orhtogonal to the mean regressor. Due to this residual-based
tests as in GUIDE break down and do not have power to detect this. Note that this situation does not
have to be rare in practice: Especially for binary regressor variables (e.g., as in treatment-subgroup
investigations) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.

@@

\section*{Computational details}
\label{sec:comp}
The applied implementation is based on the \proglang{R} package \pkg{partykit}
(version~\Sexpr{packageVersion("partykit")}) which is available 
on \proglang{R}-Forge at (\url{https://R-Forge.R-project.org/projects/partykit/})
where the code to reproduce the simulation study can be found as well.

>>>>

#if(file.exists("tree.rda")){
#  load("tree.rda")
#} else {
#  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
#                         nrsplitvar = 2, nrsteps = 1, nrlevels = 2,
#                         delta = seq(from = 0, to = 1, by = 0.1),
#                         xi = c(0, 0.2, 0.5, 0.8), vary_beta = "all",
#                         binary_regressor = FALSE, binary_beta = TRUE,
#                         only_intercept = FALSE,
#                         test = c("ctree", "mfluc", "ctree_max",
#                                  "ctree_cat", "mfluc_cat", "ctree_max_cat",
#                                  "ctree_bin", "mfluc_bin", "ctree_max_bin",
#                                  "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
#                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
#                         beta0 = NULL, beta1 = NULL,
#                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05)
#  
#  save(simres, file = "tree.rda")
#}
#sim_tree <- simres$res
#rm(simres)

@@

The top row of Figure~\ref{fig:tree_all} presents the results for pre-pruning which, not surprisingly,
reflect the results from previous sections. Thus, CTree and MOB perform similarly and rather well
as the underlying significance tests have good power properties. Only for late abrupt shifts the
linear test statistics in CTree clearly have lower power than the maximally-selected statistics in MOB.
Again, this picture would reverse for smooth changes instead of abrupt shifts. Compared to CTree/MOB
both GUIDE flavors clearly perform worse as the underlying tests are less powerful; notably
for the residual-based GUIDE which again has problems picking up the slope change associated with $Z_2$.

@@

When switching from pre-pruning to post-pruning (bottom row of Figure~\ref{fig:tree_all})
it is indeed shown that many of the problems stemming from the low power of the two GUIDE
flavors are indeed mitigated. Thus, by first growing a large tree without stopping upon
non-significance and then pruning back based on predictive performance instead substantially
improves the fit of the two GUIDE flavors. However, using the residuals only in GUIDE still
performs clearly worse compared to the GUIDE+scores. The latter is essentially on par with
CTree and MOB when the true split matches one of the categorization bins (i.e., $\xi = 0$
and $\xi = 0.5$, respectively) while CTree/MOB still perform somewhat better in the other
cases ($\xi = 0.2$ and $\xi = 0.8$).

In summary, there is clear support for the conventional wisdom that the power of the testing
strategy in an unbiased tree algorithm is not so important when combined with post-pruning.
However, there are limits to this. Consequently, when tests can be made more powerful --
e.g., by using scores instead of residuals -- then this improves the corresponding tree
algorithm. Finally, the simulation also supports using pre-pruning in an unbiased tree algorithm
when the underlying testing strategy also works well as a classical significance test.

@@

Linear models as presented in this study have been chosen as they are highly relevant in practice,
allow for simple illustrations, and theoretical insights are available for the testing strategies.
However, the results can be easily extended to a wide variety of other models where the introduced building
blocks can be applied in the same way. Specifically, it has been shown theoretically that certain
changes in the parameters do not lead to shifts in the residuals (but in other components of the
scores). \cite{Ploberger+Kraemer:1992} showed that residual-based tests can detect a change in the parameters
of a linear model only if it also causes a shift in the expected value $\mathbb{E}(Y)$. 
This is not the case if changes are orthogonal to the mean regressor which in our case is $(1,0)^{\top}$.
Consequently, if only the slope $\beta_1$ but not the intercept $\beta_0$ changes, the shift
is of type $(0, \delta)^\top$ and thus orthogonal to the mean regressor. Due to this residual-based
tests as in GUIDE break down and do not have power to detect this. Note that this situation does not
have to be rare in practice: Especially for binary regressor variables (e.g., as in treatment-subgroup
investigations) it can easily occur \citep[see Figure~2 in][for an illustration]{Loh+He+Man:2015}.

@@

\section*{Computational details}
\label{sec:comp}
The applied implementation is based on the \proglang{R} package \pkg{partykit}
(version~\Sexpr{packageVersion("partykit")}) which is available 
on \proglang{R}-Forge at \url{https://R-Forge.R-project.org/projects/partykit/}.
The code to reproduce the simulation study is available in the supplement for this 
paper on arXiv.org E-Print Archive (\url{https://arxiv.org/}).
